Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-29 paper/main.tex
paper

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such Bayesian optimization
algorithms to the combinatorial case, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets.

@@

The \textbf{LOP} is defined as 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n-1} \sum_{j=i+1}^n  a_{\sigma(i), \sigma(j)}
\end{equation}
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

For the LOP, we generated synthetic instances as follows: sample 200 permutations from a MM of size $n=20$ and $\theta=\{-\log(.5), -\log(.7), -\log(.9)\}$, i.e., from peaked distributions to almost uniform distributions. Then, generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}(\sigma(i)<\sigma(j))$.

@@

\subsubsection{Computing environment.}
All experiments were run on Intel Xeon ``Ivybridge'' E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7.4.1708 (Core).
\MANUEL{This may need updating since you ran some experiments? didn't you?}
\ekhine{}{esto es lo q se: Hipatia (Cluster)
18 nodes (1 with Nvidia Tesla K40 GPU)
624 cores
4TB of RAM Memory
Infiniband network}

>>>>

In many practical optimization problems, the objective function is not
explicitly available and solutions are evaluated by means of expensive
prediction models, simulations or physical experiments. When decision variables
are continuous, the use of Bayesian surrogate-models (e.g., Gaussian processes)
is widespread in
optimization~\citep{JonSchWel98go,ForKea2009surrogate}. Motivated by this
success, there have been attempts at adapting such Bayesian optimization
algorithms to the combinatorial case, a notable example being Combinatorial
Efficient Global Optimization
(CEGO)~\citep{ZaeStoBar2014:ppsn,ZaeStoFriFisNauBar2014}. However, the
ruggedness of combinatorial landscapes, which makes local search particularly
effective in the white-box context, lessens the effectiveness of global
surrogate models~\citep{EriPeaGar2019scalable}. Moreover, surrogate models are
expensive to train and optimize. The additional time required by the Bayesian
optimizer may impose a significant overhead in computation time, specially if
computation time of each function evaluation is measured in ``few'' minutes or
hours rather than days and when ``expensive'' refers to resources or economical
cost rather than wall-clock time. \citet{PerLopStu2015si} recently showed that
ant colony optimization (ACO) %~\citep{DorStu2004:book}
is competitive with CEGO
on a black-box version of the travelling salesman problem under a budget of
$1\,000$ function evaluations.  On the other hand, $1\,000$ evaluations is
still a relatively large budget for CEGO, whereas ACO was not designed for such
short budgets. Finally, we note that learning Budget-constrained Combinatorial is a hot topic for a variety of structures besides permutation~\cite{manchanda2020}.

@@

The \emph{Linear Ordering Problem (LOP)} is defined as 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n-1} \sum_{j=i+1}^n  a_{\sigma(i), \sigma(j)}
\end{equation}
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

For the LOP, we generated synthetic instances as follows: sample 200 permutations from a MM of size $n=20$ and $\theta=-\log(uniformity)$, where  $uniformity=\{.5,.7,.9)\}$, i.e., from peaked distributions to almost uniform distributions. Then, generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}(\sigma(i)<\sigma(j))$.

@@

\subsubsection{Computing environment.}
All experiments were run on 
Intel Xeon ``Ivybridge'' E5-2650v2 CPUs at 2.60\,GHz, 64\,GB RAM running CentOS Linux release 7.4.1708 (Core),
\ekhine{}{
Intel Xeon ``GenuineIntel'' E5-2680v3 CPUs at 2.50\,GHz, 32???\,GB RAM running Linux,
Intel Xeon ``GenuineIntel'' E5-2683v4 CPUs at 2.10\,GHz, 32???\,GB RAM running Linux.
}
%processor : 23
%vendor_id : GenuineIntel
%cpu family : 6
%model : 63
%model name : Intel(R) Xeon(R) CPU E5-2680 v3 @ 2.50GHz
%stepping : 2
%microcode : 0x43
%cpu MHz : 1199.951
%cache size : 30720 KB
%physical id : 1
%siblings : 12
%core id : 13
%cpu cores : 12
%apicid : 58
%initial apicid : 58
%fpu : yes
%fpu_exception : yes
%cpuid level : 15
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm epb intel_ppin ssbd ibrs ibpb stibp tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 avx2 smep bmi2 erms invpcid cqm xsaveopt cqm_llc cqm_occup_llc dtherm ida arat pln pts md_clear spec_ctrl intel_stibp flush_l1d
%bogomips : 5005.34
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual
%
%The node n015 has 32 processors. This is the info of the 32st processor:
%processor : 31
%vendor_id : GenuineIntel
%cpu family : 6
%model : 79
%model name : Intel(R) Xeon(R) CPU E5-2683 v4 @ 2.10GHz
%stepping : 1
%microcode : 0xb000022
%cpu MHz : 2442.608
%cache size : 40960 KB
%physical id : 1
%siblings : 16
%core id : 15
%cpu cores : 16
%apicid : 62
%initial apicid : 62
%fpu : yes
%fpu_exception : yes
%cpuid level : 20
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm 3dnowprefetch epb cat_l3 cdp_l3 intel_ppin intel_pt tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 hle avx2 smep bmi2 erms invpcid rtm cqm rdt_a rdseed adx smap xsaveopt cqm_llc cqm_occup_llc cqm_mbm_total cqm_mbm_local dtherm ida arat pln pts
%bogomips : 4207.67
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual
